---
title: "tidying"
format: html
editor: visual
---

```{r}
#|message: false

library(dplyr)
library(tidyr)
library(readr)
```

# Data Sources

## Read in our data

```{r}
#|message: false

catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

## Explore
```{r}
#print column names
colnames(catch_original)

#shows first six lines
head(catch_original)

#summary statistics of each column
summary(catch_original)

#unique values in columns
unique(catch_original)

```

# Cleaning and Reshaping

## Using the select() function

```{r}
catch_data <- catch_original %>% 
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)
head(catch_data)

catch_data <- catch_original %>% 
  select(-All, -notesRegCode)
head(catch_data)

#Chinook is a character string

```

## check column types

```{r}
glimpse(catch_data)

```
## Replace values in a column with mutate()

```{r}
#change Chinook column to numeric

catch_clean <- catch_data %>% 
  mutate(Chinook = as.numeric(Chinook))

```

```{r}
#find na values
is.na(catch_clean$Chinook)

i <- which(is.na(catch_clean$Chinook))
i

catch_data[i,]

catch_clean <- catch_data %>% 
  mutate(Chinook=if_else(condition= Chinook == "I", 
                         true = "1",
                         false = Chinook),
    Chinook = as.numeric(Chinook))


```

## Changing shape using pivot_longer() and pivot_wider()

```{r}
head(catch_clean)

#pulling 

catch_long <- catch_clean %>% 
  
  pivot_longer(cols= -c(Region, Year),
             names_to = "species", 
             values_to = "catch")

head(catch_long)

```

```{r}
catch_wide <- catch_long %>%
    pivot_wider(names_from = species,
                values_from = catch)

head(catch_wide)
```

## Renaming columns with rename()
```{r}
catch_long <- catch_long %>%
    rename(catch_thousands = catch)

head(catch_long)
```


```{r}
#changing numeric value using arithmetic operation to better understand data

catch_long <- catch_long %>%
    mutate(catch = catch_thousands*1000) %>% 
    select(-catch_thousands)

head(catch_long)
```

#summary statistics using group_by() and summarize()
```{r}
#get avg catch by region
mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarise(mean_catch=mean(catch))
head(mean_region)

#number of rows for each group
n_region <- catch_long %>% 
  group_by(Region) %>% 
  summarise(n=n())
head(n_region)

```

## Filtering rows using filter()

```{r}
sse_catch <- catch_long %>% 
  filter(Region=="SSE")
head(sse_catch)

#filter to where catches were ovver 1 million fish

catch_million <- catch_long %>% 
  filter(catch>1000000)
head(catch_million)

#filter to just Chinook for SSE
chinook_sse <- catch_long %>% 
  filter(Region=="SSE" & species=="Chinook")
head(chinook_sse)

```

## Sorting data with arrange()

```{r}
mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarise(mean_catch = mean(catch)) %>% 
  arrange(mean_catch)
head(mean_region)

#if you want number descending

mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarise(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))
head(mean_region)
```

##split and combine values in columns
```{r}
#create new dataframe

sites_df <- data.frame(site=c("HAW-101",
                               "HAW-103", 
                              "OAH-320"))

#splitting the values in one column into two
#"-" will vary, could be space, comma, etc. 

split_df <- sites_df %>% 
  separate(site, c("island", "site_number"), "-")


```

```{r}
dates_df <- data_frame(
  year = c("1930", "1930", "1930"),
  month = c("12","12", "12"),
  day = c("14","15","16"))

dates_df

dates_unite <- dates_df %>% 
  unite(date, year, month, day, sep = "-")
head(dates_unite)
```

## now all together

```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")

mean_region <- catch_original %>% 
  select(-All, -notesRegCode) %>% 
  mutate(Chinook=if_else(condition= Chinook == "I", 
                         true = "1",
                         false = Chinook),
    Chinook = as.numeric(Chinook)) %>% 
pivot_longer(cols= -c(Region, Year),
            names_to = "species", 
            values_to = "catch") %>% 
  mutate(catch = catch*1000) %>% 
  group_by(Region) %>% 
  summarise(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))

head(mean_region)

```
## Save your results with write_csv()

```{r}

#creates file

write_csv(mean_region, here::here("data/mean_catch_by_region.csv"))

#data/ is file directory
```




